---
title: "Using YZFishDB with regionbarcoder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using YZFishDB with regionbarcoder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This vignette demonstrates the first `regionbarcoder` workflow using YZFishDB,
a curated mitochondrial DNA reference database for fishes of the Yangtze River
Basin. The workflow connects to a SQLite release, inspects available tables,
queries marker-specific references, exports pipeline-ready FASTA files, and
summarizes database diagnostics.

## Connect to a reference database

```{r}
library(regionbarcoder)

con <- rb_connect()
rb_tables(con)
```

By default, `rb_connect()` uses the bundled demonstration database when present.
During development in this project folder, it can also locate the full local
`zenodo/data/YZFishDB.db` release.

For production use, the full YZFishDB SQLite release is archived outside the R
package on Zenodo: <https://doi.org/10.5281/zenodo.18155084>. Users can then
download and cache it once:

```{r, eval = FALSE}
rb_install_db()
con <- rb_connect()
```

`rb_install_db()` uses the Zenodo-hosted `YZFishDB.db` file by default and checks
the expected file size and md5 checksum. The release metadata used by the
package can be inspected with:

```{r, eval = FALSE}
rb_yzfishdb_release()
```

The full SQLite file is about 606 MB. On slow networks, increase the timeout
and retry count:

```{r, eval = FALSE}
rb_install_db(timeout = 7200, retries = 5)
```

If the download is interrupted repeatedly, users can download `YZFishDB.db`
manually from the Zenodo DOI page and set `RB_YZFISHDB_PATH` to that local file.

If a mirror or future release URL is configured in the environment,
`rb_connect()` can install the database automatically when the local cache is
missing:

```{r, eval = FALSE}
Sys.setenv(RB_YZFISHDB_URL = "https://example.org/YZFishDB.db")

con <- rb_connect(download = TRUE)
```

## Query native 12S references

```{r}
native_12s <- rb_get_sequences(con, marker = "12S", occurrence = "native")
nrow(native_12s)
head(native_12s[, c("source", "sequence_id", "species", "seq_type", "family")])
```

The same interface can subset by species, genus, family, source, occurrence,
and QC status. This makes database filtering explicit and reproducible.

## Export pipeline-ready references

```{r}
example_refs <- head(native_12s, 5)

dada2_file <- tempfile(fileext = ".fasta")
qiime2_fasta <- tempfile(fileext = ".fasta")
qiime2_tax <- tempfile(fileext = ".tsv")
blast_file <- tempfile(fileext = ".fasta")
usearch_file <- tempfile(fileext = ".fasta")

rb_export_dada2(example_refs, dada2_file)
rb_export_qiime2(example_refs, qiime2_fasta, qiime2_tax)
rb_export_blastn(example_refs, blast_file)
rb_export_usearch(example_refs, usearch_file)

readLines(dada2_file, n = 4)
readLines(qiime2_tax, n = 3)
```

These exports preserve a consistent seven-rank taxonomy string while adapting
headers to DADA2, QIIME2, BLASTN, and USEARCH/VSEARCH conventions.

## Assign eDNA ASVs against YZFishDB

Most users will generate ASVs with DADA2 or QIIME2 first, then use
`regionbarcoder` for taxonomic assignment against YZFishDB. The package supports
a BLASTN workflow when BLAST+ is installed. For a dependency-free demonstration,
the example below uses exact matching with one sequence taken from the reference
database.

```{r}
example_asv <- tempfile(fileext = ".fasta")
writeLines(c(">ASV_001", example_refs$sequence[[1]]), example_asv)

assignment <- rb_assign_edna(
  asv_fasta = example_asv,
  con = con,
  marker = "12S",
  method = "exact"
)

assignment
```

If the user has an ASV count table, assignments can be converted to a
sample-by-species community matrix.

```{r}
asv_counts <- data.frame(
  asv_id = "ASV_001",
  sample_a = 12,
  sample_b = 3,
  check.names = FALSE
)

rb_build_species_matrix(assignment, asv_counts)
```

## Summarize coverage and QC outcomes

```{r}
coverage <- rb_marker_coverage(con)
head(coverage)

sources <- rb_source_coverage(con)
head(sources)

qc <- rb_qc_summary(con)
head(qc)
```

These summaries are designed for reporting the reference database before
assignment results are interpreted.

## Barcode gap and ambiguity tables

```{r}
barcode <- rb_barcode_gap(con)
head(barcode)

ambiguous <- rb_ambiguity(con)
head(ambiguous)
```

The full YZFishDB release stores barcode gap and ambiguity outputs as external
tables in the Zenodo data package. A future release can merge these directly
into the SQLite release so `regionbarcoder` can report them without additional
CSV imports.

```{r}
rb_disconnect(con)
```
