---
title: "Case study: Taxonomic assignment of Yangtze River eDNA using YZFishDB"
author: "Kua Zi Xun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Case study: Yangtze River eDNA}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, tidy = TRUE, collapse = TRUE, comment = "#>")
Sys.setenv(RB_YZFISHDB_PATH = "C:/Zi Xun/IHB/Writing/kzx_etal_refdb/zenodo/data/YZFishDB.db")
```

# Introduction

This vignette demonstrates the complete workflow of the `regionbarcoder` package using a real-world environmental DNA (eDNA) dataset from the Yangtze River basin. We will walk through auditing the `YZFishDB` reference database, benchmarking assignment methods (Exact vs. BLASTn), resolving taxonomic ties, and generating a final species-by-site ecological matrix.

*Note: This case study utilizes the `YZFishDB.db` reference database and unpublished ASV data. To run this code locally, ensure you update the file paths in the setup chunk to match your system.*

# 1. Setup and Configuration

First, we load the required libraries and define the paths to our reference database, ASV data, and BLAST+ executables.

```{r setup, message=FALSE, warning=FALSE}
library(regionbarcoder)
library(tidyverse)
library(Biostrings)

# --- USER CONFIGURATION: Update these paths for your local machine ---
db_path       <- "path/to/YZFishDB.db"
asv_path      <- "path/to/novogene_202425_edna.rds"
site_path     <- "path/to/yz_site_info_tbl_2025.csv"
blastn_path   <- "path/to/blastn.exe"
makeblastdb_p <- "path/to/makeblastdb.exe"

# Output directories
dir.create("data", showWarnings = FALSE)
asv_fasta     <- "data/novogene_asvs.fasta"
blast_out_dir <- "data/regionbarcoder_blastout"
```

# 2. Auditing the Reference Database

Before assigning sequences, it is crucial to understand the composition of the reference database. `regionbarcoder` provides built-in functions to audit taxonomic and marker coverage dynamically.

```{r audit-db}
con <- rb_connect(db_path)

# 1. Audit QC Flags
# (Note: yzfishdb_final primarily contains 'pass' sequences)
qc_summary <- rb_qc_summary(con)
print(qc_summary)

# 2. Compare Taxonomic Coverage (12S vs. Full Database)
species_12s <- rb_list_taxa(con, rank = "species", marker = "12S")
species_all <- rb_list_taxa(con, rank = "species")

cat("Species with 12S data:", nrow(species_12s), "\n")
cat("Species with any marker data:", nrow(species_all), "\n")

# 3. Source Coverage
source_cov <- rb_source_coverage(con)
print(head(source_cov))
```

# Visualizing Source Coverage

We can visualize the provenance of the reference sequences to understand the composition of the database across different markers.

```{r}
source_cov_plot_data <- source_cov %>%
  mutate(Source = factor(source, 
                         levels = c("mitofish", "bold", "ncbi", "midori", "local", "mitofish|ncbi"),
                         labels = c("MitoFish", "BOLD", "NCBI", "MIDORI2", "Local", "MitoFish & NCBI")))

p1 <- ggplot(source_cov_plot_data, aes(x = Source, y = n_sequences, fill = seq_type)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set1", name = "Marker") +
  labs(y = "Sequence Count", title = "Absolute Sequence Counts by Source") +
  theme_classic() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(source_cov_plot_data, aes(x = Source, y = n_sequences, fill = seq_type)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_brewer(palette = "Set1", name = "Marker") +
  labs(y = "Proportion", title = "Proportion of Markers by Source") +
  theme_classic() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine and save
ggpubr::ggarrange(p1, p2, common.legend = TRUE, ncol = 2)
```

# 3. Preparing ASV Data

The package requires ASV sequences in a FASTA file and a corresponding count table. We extract these from a standard DADA2 sequence table (`seqtab`).

```{r}
seqtab <- readRDS(asv_path)

# Extract sequences and assign standardized ASV IDs
asv_seqs <- colnames(seqtab)
asv_ids <- sprintf("ASV_%05d", seq_along(asv_seqs))
names(asv_seqs) <- asv_ids

# Write to FASTA
writeXStringSet(DNAStringSet(asv_seqs), filepath = asv_fasta)

# Create ASV count table (Samples x ASVs)
asv_table <- as.data.frame(t(seqtab))
asv_table$asv_id <- asv_ids
asv_table <- asv_table[, c("asv_id", setdiff(names(asv_table), "asv_id"))]
```

# 4. Taxonomic Assignment: Exact vs. BLASTn

A key feature of `regionbarcoder` is the ability to easily benchmark different assignment methods. Here, we compare the dependency-free exact matching method against the robust BLASTn method.

```{r}
# Run EXACT Match
t_exact <- system.time({
  res_exact <- rb_assign_edna(
    asv_fasta = asv_fasta, con = con, marker = "12S", 
    method = "exact", out_dir = blast_out_dir
  )
})

# Run BLASTn
t_blast <- system.time({
  res_blast <- rb_assign_edna(
    asv_fasta = asv_fasta, con = con, marker = "12S", 
    method = "blastn", out_dir = blast_out_dir,
    blastn = blastn_path, makeblastdb = makeblastdb_p
  )
})

# Compare Results
comparison <- data.frame(
  Method = c("Exact", "BLASTn"),
  Time_seconds = c(t_exact[3], t_blast[3]),
  ASVs_Assigned = c(
    sum(res_exact$assignment_status != "no_match"),
    sum(res_blast$assignment_status != "no_match")
  ),
  Success_Rate_Pct = c(
    round(mean(res_exact$assignment_status != "no_match") * 100, 2),
    round(mean(res_blast$assignment_status != "no_match") * 100, 2)
  )
)
print(comparison)
```

*Note: As expected, Exact matching is significantly faster but yields a lower success rate due to its strict 100% identity requirement. BLASTn is the recommended method for complex environmental samples.*

# 5. Resolving Ambiguities (Manual Curation)

The upgraded BLASTn pipeline uses advanced metrics (`ccovs`, `bitscore_pb`) and automatically flags perfect ties as "`check_full_tie`". This allows researchers to easily export and manually curate ambiguous ASVs based on biogeography or taxonomy.

```{r}
# Inspect Ties
ties <- res_blast %>% filter(assignment_status == "check_full_tie")
cat("Number of ASVs requiring manual curation:", nrow(ties), "\n")

if(nrow(ties) > 0) {
  print(arrange(ties[, c("asv_id", "tie_taxa", "ccovs", "bitscore_pb")], tie_taxa, asv_id))
}

# --- Mock Manual Curation ---
# Example: ASV_00208 is a tie between Macrognathus aculeatus and Sinobdella sinensis.
# Based on literature, S. sinensis is the correct assignment for this region.
if("ASV_00208" %in% ties$asv_id) {
  res_blast$species[res_blast$asv_id == "ASV_00208"] <- "Sinobdella sinensis"
  res_blast$assignment_status[res_blast$asv_id == "ASV_00208"] <- "unique"
  res_blast$tie_taxa[res_blast$asv_id == "ASV_00208"] <- NA_character_
}
```

# 6. Building the Ecological Matrix

Finally, we aggregate the curated ASV counts into a species-by-sample matrix, dropping any remaining ambiguous or unassigned ASVs.

```{r}
species_sample <- rb_build_species_matrix(
  assignment = res_blast,
  asv_table = asv_table,
  include_ambiguous = FALSE, 
  unassigned = FALSE         
)

cat("Final ecological matrix dimensions (Species x Samples):", dim(species_sample), "\n")
head(species_sample)
```

## Mapping to Site Metadata and Visualization

To visualize the data, we map the raw sample IDs to their corresponding site IDs and generate a heatmap.

```{r}
site_info <- read.csv(site_path) %>% mutate(sample_id = toupper(sample_id))

# Pivot and join with site metadata
species_matrix_long <- species_sample %>%
  as.data.frame() %>%
  rownames_to_column("species") %>%
  pivot_longer(-species, names_to = "sample_id", values_to = "abundance") %>%
  mutate(sample_id = toupper(sample_id)) %>%
  inner_join(dplyr::select(site_info, site_id, sample_id), by = "sample_id") %>%
  filter(abundance > 0)

# Pivot back to wide format for heatmap
species_matrix <- species_matrix_long %>%
  pivot_wider(names_from = site_id, values_from = abundance, values_fill = 0) %>%
  column_to_rownames('species') %>%
  as.matrix()

# Generate and save high-resolution heatmap
png("figures/species_site_heatmap.png", width = 11, height = 10, units = "in", res = 400)
heatmap(species_matrix, Colv = NA, Rowv = NA, scale = 'column', 
        revC = TRUE, xlab = "Sites", cexRow = 0.5, cexCol = 0.7,
        main = "Species Abundance Across Yangtze River Sites")
dev.off()
```

# Conclusion

This case study demonstrates how `regionbarcoder` streamlines the eDNA taxonomic assignment workflow. By dynamically querying the `YZFishDB` database, utilizing robust alignment metrics (`ccovs`), and providing a standardized framework for tie curation (`check_full_tie`), the package ensures reproducible, transparent, and bioinformatically rigorous species assignments.

```{r}
rb_disconnect(con)
```